A Two-Infall Star Formation History with Outflows and Radial Migration

### 2025-01-31 src/tex/ms.tex
Updates from Overleaf

\begin{table*}
    \centering
    \caption{Sample selection parameters and median uncertainties from APOGEE DR17 (see Section \ref{sec:observational-sample}).}
    \label{tab:sample}
    \begin{tabular}{lll}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Range or Value} & \multicolumn{1}{c}{Notes} \\
        \hline
        $\log g$            & $1.0 < \log g < 3.8$          & Select giants only \\
        $T_{\rm eff}$       & $3500 < T_{\rm eff} < 5500$ K & Reliable temperature range \\
        $S/N$               & $S/N > 80$                    & Required for accurate stellar parameters \\
        ASPCAPFLAG Bits     & $\notin$ 23                   & Remove stars flagged as bad \\
        EXTRATARG Bits      & $\notin$ 0, 1, 2, 3, or 4     & Select main red star sample only \\
        Age                 & $\sigma_{\rm Age} < 40\%$     & Age uncertainty from \citetalias{leung_variational_2023} \\
        \hline
        $\sigma({\rm [Fe/H]})$ & $9.2\times10^{-3}$ & Median uncertainty in [Fe/H] \\
        $\sigma({\rm [O/Fe]})$ & $1.8\times10^{-2}$ & Median uncertainty in [O/Fe] \\
        $\sigma($log(Age/Gyr)) & $0.10$ & Median age uncertainty \citepalias{leung_variational_2023} \\
        \hline
    \end{tabular}
\end{table*}

@@

We restrict our sample to red giant branch and red clump stars with high-quality spectra. Table \ref{tab:sample} lists our selection criteria, which largely follow from \citet{Hayden2015-ChemicalCartography}. This produces a final sample of \input{output/sample_size.txt}stars with [O/Fe] and [Fe/H] abundance measurements. APOGEE stars are matched with their Bailer-Jones photo-geometric distance estimate from \textit{Gaia} Early Data Release 3 \citep{gaia_collaboration_gaia_2016,gaia_collaboration_gaia_2021}, which we use to calculate galactocentric radius $R_{\rm gal}$ and midplane distance $z$.

We use estimated ages from \citet[][hereafter \citetalias{leung_variational_2023}]{leung_variational_2023}, who use a variational encoder-decoder network which is trained on asteroseismic data to retrieve age estimates for APOGEE giants without contamination from age-abundance correlations. Importantly, the \citetalias{leung_variational_2023} ages do not plateau beyond $\sim10$ Gyr as they do in astroNN \citep{Mackereth2019-astroNN-Ages}. We use an age uncertainty cut of 40\% per the recommendations of \citetalias{leung_variational_2023}, which produces a total sample of \input{output/age_sample_size.txt}APOGEE stars with age estimates. We note that we use the full sample of \input{output/sample_size.txt}APOGEE stars unless we explicitly compare to age estimates. Table \ref{tab:sample} also presents the median uncertainties in the data for [Fe/H], [O/Fe], and age.

@@

\begin{figure*}
    \centering
    \includegraphics{src/tex/figures/ofe_df_comparison.pdf}
    \caption{Stellar [O/Fe] distributions produced by multi-zone models (a--d) and as observed by APOGEE (d). Stars are binned by Galactocentric radius, represented by the color scale, and absolute midplane distance $|z|$, represented by the different rows. Each distribution is normalized so that the area under the curve is 1, and the vertical scale is consistent across each row. The median APOGEE abundance uncertainties were forward-modeled onto the model outputs. For visual clarity, each distribution is smoothed with a box-car of width 0.05 dex. Model (a) assumes the fiducial parameters and the {\tt yZ1} yield set, whereas models (b--d) assume the {\tt yZ2} yield set. Model (c) also assumes a power-law SN Ia DTD, while model (d) assumes the infalling gas has metallicity ${\rm [O/H]}_{\rm CGM}={\rm [Fe/H]}_{\rm CGM}=-0.7$. The distributions from APOGEE DR17 are plotted in column (e) for reference. {\bf Key takeaway:} The high-yield case produces a peak at intermediate [O/Fe] which is prominent in most galactic regions, which can be mitigated by pre-enriched infall.}
    \label{fig:ofe-df}
\end{figure*}

As noted by \citet{dubay_galactic_2024} two-infall model generically predicts {\it three} peaks in the [O/Fe] distribution which correspond to the high-$\alpha$ sequence, the abundance ``turn-over'' after the second infall, and finally the late-time low-$\alpha$ sequence. Figure \ref{fig:ofe-df} compares [O/Fe] distributions across the Galactic disk produced by four different multi-zone models with the distributions observed in APOGEE DR17. For the fiducial model (a) with $y/Z_\odot=1$, the latter two peaks are close together and the overall effect is similar to the observed distribution. With a higher yield set, however, there is a $\sim0.2$ dex separation between the low- and intermediate-$\alpha$ peaks, because the CCSN element production after the second infall is more efficient. As a result, model (b) predicts a high density of stars in the APOGEE ``trough.''

>>>>

\begin{table*}
    \centering
    \caption{Sample selection parameters and median uncertainties from APOGEE DR17 (see Section \ref{sec:observational-sample}).}
    \label{tab:sample}
    \begin{tabular}{lll}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Range or Value} & \multicolumn{1}{c}{Notes} \\
        \hline
        $\log g$            & $1.0 < \log g < 3.8$          & Select giants only \\
        $T_{\rm eff}$       & $3500 < T_{\rm eff} < 5500$ K & Reliable temperature range \\
        $S/N$               & $S/N > 80$                    & Required for accurate stellar parameters \\
        ASPCAPFLAG Bits     & $\notin$ 23                   & Remove stars flagged as bad \\
        EXTRATARG Bits      & $\notin$ 0, 1, 2, 3, or 4     & Select main red star sample only \\
        Age                 & $\sigma_{\rm Age} < 40\%$     & Age uncertainty from \citetalias{Leung2023-Ages} \\
        $R_{\rm gal}$     & $3 < R_{\rm gal} < 15$ kpc    & Eliminate bulge \& extreme outer-disk stars \\
        $|z|$               & $|z| < 2$ kpc                 & Eliminate halo stars \\
        \hline
    \end{tabular}
\end{table*}

\begin{table*}
    \centering
    \caption{Median and dispersion in APOGEE parameter uncertainties.}
    \label{tab:uncertainties}
    \begin{tabular}{lcc}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Median Uncertainty} & \multicolumn{1}{c}{Uncertainty Dispersion ($95\%-5\%$)} \\
        \hline
        [Fe/H]          & $0.0089$   & $0.0060$ \\
        ${\rm [O/Fe]}$  & $0.019$    & $0.031$ \\
        log(Age/Gyr)    & $0.10$     & $0.16$ \\
        \hline
    \end{tabular}
\end{table*}

@@

We restrict our sample to red giant branch and red clump stars with high-quality spectra. Table \ref{tab:sample} lists our selection criteria, which largely follow from \citet{hayden_chemical_2015}. This produces a final sample of \input{output/sample_size.txt} stars with calibrated [O/Fe] and [Fe/H] abundance measurements. We make use of the {\it Gaia} Early Data Release 3 (EDR3) data \citep{gaia_collaboration_gaia_2016,gaia_collaboration_gaia_2021} included in the catalog by the APOGEE team. Specifically, we use the \citet{bailer-jones_estimating_2021} photo-geometric distance estimates to calculate galactocentric radius $R_{\rm gal}$ and midplane distance $z$, assuming a Sun--Galactic center distance $R_\odot=8.122$ kpc \citep{gravity_collaboration_detection_2018} and height of the Sun above the midplane $z_\odot=20.8$ pc \citep{bennett_vertical_2019}. Table \ref{tab:apogee-regions} lists the number of APOGEE stars in bins of $R_{\rm gal}$ and $|z|$. For some Galactic regions with $R_{\rm gal}<5$ kpc or $R_{\rm gal}>13$ kpc, the median distance error exceeds 1 kpc but remains within our bin width of 2 kpc, and the vast majority of stars have much smaller distance uncertainties.

We use estimated ages from \citet[][hereafter \citetalias{leung_variational_2023}]{leung_variational_2023}, who use a variational encoder-decoder network which is trained on asteroseismic data to retrieve age estimates for APOGEE giants without contamination from age-abundance correlations. Importantly, the \citetalias{leung_variational_2023} ages do not plateau beyond $\sim10$ Gyr as they do in astroNN \citep{mackereth_dynamical_2019}. We use an age uncertainty cut of 40\% per the recommendations of \citetalias{leung_variational_2023}, which produces a total sample of \input{output/age_sample_size.txt} APOGEE stars with age estimates. We note that we use the full sample of \input{output/sample_size.txt} APOGEE stars unless we explicitly compare to age estimates. Table \ref{tab:uncertainties} presents the median and dispersion ($95^{\rm th} - 5^{\rm th}$ percentile difference) of the uncertainty in [Fe/H], [O/Fe], and log(age).

@@

\begin{figure*}
    \centering
    \includegraphics{src/tex/figures/ofe_df_comparison.pdf}
    \caption{Stellar [O/Fe] distributions produced by multi-zone models (a--d) and as observed by APOGEE (d). Stars are binned by Galactocentric radius, represented by the color scale, and absolute midplane distance $|z|$, represented by the different rows. Each distribution is normalized so that the area under the curve is 1, and the vertical scale is consistent across each row. The median APOGEE abundance uncertainties were forward-modeled onto the model outputs. For visual clarity, each distribution is smoothed with a box-car of width 0.05 dex. Model (a) assumes the fiducial parameters and the {\tt yZ1} yield set, whereas models (b--d) assume the {\tt yZ2} yield set. Model (c) also assumes a power-law SN Ia DTD, while model (d) assumes the infalling gas has metallicity ${\rm [O/H]}_{\rm CGM}={\rm [Fe/H]}_{\rm CGM}=-0.7$. The distributions from APOGEE DR17 are plotted in column (e) for reference.}
    \label{fig:ofe-df}
\end{figure*}

As discussed in Section \ref{sec:sfh}, the two-infall model generically predicts {\it three} peaks in the [O/Fe] distribution which correspond to the high-$\alpha$ sequence, the abundance ``turn-over'' after the second infall, and finally the late-time low-$\alpha$ sequence. Figure \ref{fig:ofe-df} compares [O/Fe] distributions across the Galactic disk produced by four different multi-zone models with the distributions observed in APOGEE DR17. For the fiducial model (a) with $y/Z_\odot=1$, the latter two peaks are close together and the overall effect is similar to the observed distribution. With a higher yield set, however, there is a $\sim0.2$ dex separation between the low- and intermediate-$\alpha$ peaks, because the CCSN element production after the second infall is more efficient. As a result, model (b) predicts a high density of stars in the APOGEE ``trough.''

In columns (c) and (d) of Figure \ref{fig:ofe-df}, we attempt to mitigate the issue of the intermediate-$\alpha$ peak for the $y/Z_\odot=2$ yield set. Model (c) switches out our fiducial DTD with a power-law DTD, which decreases the timescale for Fe production (see Figure \ref{fig:dtd}). However, as shown by \citet{dubay_galactic_2024}, models with a power-law DTD struggle to reproduce the high-$\alpha$ sequence for a wide range of star formation histories, and the results are no different for this two-infall model. Finally, in model (d) the metallicity of the infalling gas increases to ${\rm [X/H]}_{\rm CGM}=-0.7$ at late times, which results in very similar [O/Fe] distributions to the $y/Z_\odot=1$ case. We assume that the infalling gas has ${\rm [O/Fe]}=0$ at all times; an alternate run with ${\rm [O/Fe]}=+0.3$ shifted the distribution towards higher [O/Fe], worsening agreement with observations. In summary, pre-enriched gas infall may be necessary for the two-infall model to match the observed [O/Fe] distribution across the disk.
